Programming Language Foundations in Agda

### 2019-07-15 papers/scp/PLFA.tex
changes to intro and conclusion

\begin{abstract}
  One of the leading textbooks for formal methods is \emph{Software Foundations} (SF), written by Benjamin Pierce in collaboration with others, and based on Coq. After five years using SF in the classroom, I have come to the conclusion that Coq is not the best vehicle for this purpose, as too much of the course needs to focus on learning tactics for proof derivation, to the cost of learning programming language theory.  Accordingly, I have written a new textbook, \emph{Programming Language Foundations in Agda} (PLFA). PLFA covers much of the same ground as SF, although it is not a slavish imitation.

  What did I learn from writing PLFA? First, that it is possible. One might expect that without proof tactics that the proofs become too long, but in fact proofs in PLFA are about the same length as those in SF. Proofs in Coq require an interactive environment to be understood, while proofs in Agda can be read on the page.  Second, that constructive proofs of preservation and progress give immediate rise to a prototype evaluator. This fact is obvious in retrospect but it is not exploited in SF (which instead provides a separate normalise tactic) nor can I find it in the literature.  Third, that using raw terms with a separate typing relation is far less perspicuous than using inherently-typed terms. SF uses the former presentation, while PLFA presents both; the former uses about 1.6 as many lines of Agda code as the latter, roughly the golden ratio.

@@

Since 2013, I have taught a course on Types and Semantics for
Programming Languages to fourth-year undergraduates and masters
students at the University of Edinburgh.  An earlier version of that
course was based on \emph{Types and Programming Languages} by
\citet{Pierce-2002}, but my version was taught from its successor,
\emph{Software Foundations} (hence, SF) by \citet{Pierce-et-al-2010},
which is based on the proof assistance Coq \citep{Huet-et-al-1997}.
I am convinced by the claim of \citet{Pierce-2009}, made in his ICFP
Keynote \emph{Lambda, The Ultimate TA}, that basing a course around a
proof assistant aids learning.

However, after five years of experience, I have come to the conclusion
that Coq is not the best vehicle.  Too much of the course needs to
focus on learning tactics for proof derivation, to the cost of
learning the fundamentals of programming language theory.  Every
concept has to be learned twice: e.g., both the product data type, and
the corresponding tactics for introduction and elimination of
conjunctions.  The rules Coq applies to generate induction hypotheses
can sometimes seem mysterious.  While the \texttt{notation} construct
permits pleasingly flexible syntax, it can be confusing that the same
concept must always be given two names, e.g., both
\texttt{subst~N~x~M} and \texttt{N~[x~:=~M]}.  Names of tactics are
sometimes short and sometimes long; naming conventions in the standard
library can be wildly inconsistent.  \emph{Propositions as types} as a
foundation of proof is present but hidden.

I found myself keen to recast the course in Agda \citep{Bove-et-al-2009}.
In Agda, there is
no longer any need to learn about tactics: there is just
dependently-typed programming, plain and simple. Introduction is
always by a constructor, elimination is always by pattern
matching. Induction is no longer a mysterious separate concept, but
corresponds to the familiar notion of recursion. Mixfix syntax is
flexible while using just one name for each concept, e.g.,
substitution is \texttt{\_[\_:=\_]}. The standard library is not perfect, but
there is a fair attempt at consistency. \emph{Propositions as types} as a
foundation of proof is on proud display.

@@

The original goal was to simply adapt \emph{Software Foundations},
maintaining the same text but transposing the code from Coq to Agda.
But it quickly became clear to me that after five years in the
classroom I had my own ideas about how to present the material.  They
say you should never write a book unless you cannot \emph{not} write the
book, and I soon found that this was a book I could not not write.

I am fortunate that my student, Wen Kokke, was keen to help.  She
guided me as a newbie to Agda and provided an infrastructure that is
easy to use and produces pages that are a pleasure to view.
The bulk of the book was written January--June 2018, while on
sabbatical in Rio de Janeiro.

This paper is a personal reflection, summarising what I learned in the
course of writing the textbook. Some of it reiterates advice that is
well-known to some members of the dependently-typed programming
community, but which deserves to be better known.  The paper is
organised as follows.

@@

Section~3 compares Agda and Coq as vehicles for pedagogy.  Before
writing the book, it was not obvious that it was even possible;
conceivably, without tactics some of the proofs might
balloon in size. In fact, it turns out that for the results in PLFA
and SF, the proofs are of roughly comparable size, and (in my opinion)
the proofs in PLFA are more readable and have a pleasing visual
structure.

Section~4 observes that constructive proofs of progress and
preservation combine trivially to produce a constructive evaluator for
terms.  This idea is obvious once you have seen it, yet I cannot
find it described in the literature.  For instance, SF separately
implements a \texttt{normalise} tactic that has nothing to do with
progress and preservation.

@@

I will argue that Agda has advantages over Coq for
pedagogic purposes.  My focus is purely on the case of a proof assistant
as an aid to \emph{learning} formal semantics using examples of
\emph{modest} size.  I admit up front that
there are many tasks for which Coq is better suited than Agda.
A proof assistant that supports tactics, such as Coq or Isabelle,
is essential for formalising serious mathematics,
such as the Four-Colour Theorem \citep{Gonthier-2008},
the Odd-Order Theorem \citep{Gonthier-et-al-2013},
or Kepler's Conjecture \citep{Hales-et-al-2017},
or for establishing correctness of software at scale,
as with the CompCert compiler \citep{Leroy-2009,Kastner-et-al-2017}
or the SEL4 operating system \citep{Klein-2009,O'Connor-2016}.

@@

The book is broken into two parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
(SF is divided into books, the first two of which have the same names
as the two parts of PLFA, and cover similar material.)

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an inherently-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.

@@

There is more material that would be desirable to include in PLFA which was not
due to limits of time.  In future years, PLFA may be extended to cover
additional material, including mutable references, normalisation, System F, and
pure type systems. We'd especially like to include pure type systems as they
provide the readers with a formal model close to the dependent types used in the
book.  Our attempts so far to formalise pure type systems have proved
challenging, to say the least.

@@

There was also a pleasing surprise regarding the structure of the
proofs. While most proofs in both SF and PLFA are carried out by
induction over the evidence that a term is well typed, in SF the
central proof, that substitution preserves types, is carried out by
induction on terms for a technical reason
(the context is extended by a variable binding, and hence not
sufficiently ``generic'' to work well with Coq's induction tactic). In
Agda, we had no trouble formulating the same proof over evidence that
the term is well typed, and didn't even notice SF's description of the
issue until I was done.

@@

\section{Experience}

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  He taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty two-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA.

@@

We look forward to experience teaching from the new text,
and encourage others to use it too.  Please comment!

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing me how
much more compact it is to avoid raw terms, thanks to David Darais.  For
inspiring my work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.

>>>>

\begin{abstract}

  One of the leading textbooks for formal methods is \emph{Software
    Foundations} (SF), written by Benjamin Pierce in collaboration
  with others, and based on Coq. After five years using SF in the
  classroom, we came to the conclusion that Coq is not the best
  vehicle for this purpose, as too much of the course needs to focus
  on learning tactics for proof derivation, to the cost of learning
  programming language theory.  Accordingly, we have written a new
  textbook, \emph{Programming Language Foundations in Agda}
  (PLFA). PLFA covers much of the same ground as SF, although it is
  not a slavish imitation.

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using raw terms with a separate typing relation is far less
  perspicuous than using inherently-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

Since 2013, one of us (Philip) has taught a course on Types and Semantics for
Programming Languages to fourth-year undergraduates and masters
students at the University of Edinburgh.  An earlier version of that
course was based on \emph{Types and Programming Languages} by
\citet{Pierce-2002}, but this version was taught from its successor,
\emph{Software Foundations} (hence, SF) by \citet{Pierce-et-al-2010},
which is based on the proof assistance Coq \citep{Huet-et-al-1997}.
We are convinced by the claim of \citet{Pierce-2009}, made in his ICFP
Keynote \emph{Lambda, The Ultimate TA}, that basing a course around a
proof assistant aids learning.

However, after five years of experience, Phil came to the conclusion
that Coq is not the best vehicle.  Too much of the course needs to
focus on learning tactics for proof derivation, to the cost of
learning the fundamentals of programming language theory.  Every
concept has to be learned twice: e.g., both the product data type, and
the corresponding tactics for introduction and elimination of
conjunctions.  The rules Coq applies to generate induction hypotheses
can sometimes seem mysterious.  While the \texttt{notation} construct
permits pleasingly flexible syntax, it can be confusing that the same
concept must always be given two names, e.g., both
\texttt{subst~N~x~M} and \texttt{N~[x~:=~M]}.  Names of tactics are
sometimes short and sometimes long; naming conventions in the standard
library can be wildly inconsistent.  \emph{Propositions as types} as a
foundation of proof is present but hidden.

We found ourselves keen to recast the course in Agda \citep{Bove-et-al-2009}.
In Agda, there is
no longer any need to learn about tactics: there is just
dependently-typed programming, plain and simple. Introduction is
always by a constructor, elimination is always by pattern
matching. Induction is no longer a mysterious separate concept, but
corresponds to the familiar notion of recursion. Mixfix syntax is
flexible while using just one name for each concept, e.g.,
substitution is \texttt{\_[\_:=\_]}. The standard library is not perfect, but
there is a fair attempt at consistency. \emph{Propositions as types} as a
foundation of proof is on proud display.

@@

The original goal was to simply adapt \emph{Software Foundations},
maintaining the same text but transposing the code from Coq to Agda.
But it quickly became clear that after five years in the
classroom Philip had his own ideas about how to present the material.  They
say you should never write a book unless you cannot \emph{not} write the
book, and Philip soon found that this was a book he could not not
write.

Philip considered himself fortunate that his student, Wen, was
keen to help.  She guided Philip as a newbie to Agda and provided an
infrastructure for the book that we found easy to use and produces
pages that are a pleasure to view.  The bulk of the first draft of the
book was written January--June 2018, while Philip was on sabbatical in
Rio de Janeiro.  After the first draft was published, Jeremy wrote
two further parts.  One explains a new and simple approach
to denotational semantics, developed as part of his research.  The
other develops important precursors to the semantics, including
confluence and big-step semantics.

This paper is a personal reflection, summarising what was learned in the
course of writing the textbook. Some of it reiterates advice that is
well-known to some members of the dependently-typed programming
community, but which deserves to be better known.  The paper is
organised as follows.

@@

Section~3 compares Agda and Coq as vehicles for pedagogy.  Before
writing the book, it was not obvious that it was even possible;
conceivably, without tactics some of the proofs might
balloon in size. In fact, it turns out that for the results in PLFA
and SF, the proofs are of roughly comparable size, and (in our opinion)
the proofs in PLFA are more readable and have a pleasing visual
structure.

Section~4 observes that constructive proofs of progress and
preservation combine trivially to produce a constructive evaluator for
terms.  This idea is obvious once you have seen it, yet we cannot
find it described in the literature.

@@

Section~6 describes experience teaching from the textbook.  The point
of proof is perfection, and it turns out that an online final
examination with access to a proof assistant can lead to flawless
student performance.

Section~7 outlines our experience publishing the book as open source
in Github.  We were surprised at how effective this was at eliciting
community participation.  A large number of people have submitted pull
requests to improve the book.

We argue that Agda has advantages over Coq for
pedagogic purposes.  Our focus is purely on the case of a proof assistant
as an aid to \emph{learning} formal semantics using examples of
\emph{modest} size.  We admit up front that
there are many tasks for which Coq is better suited than Agda.
A proof assistant that supports tactics, such as Coq or Isabelle,
is essential for formalising serious mathematics,
such as the Four-Colour Theorem \citep{Gonthier-2008},
the Odd-Order Theorem \citep{Gonthier-et-al-2013},
or Kepler's Conjecture \citep{Hales-et-al-2017},
or for establishing correctness of software at scale,
as with the CompCert compiler \citep{Leroy-2009,Kastner-et-al-2017}
or the SEL4 operating system \citep{Klein-2009,O'Connor-2016}.

@@

The book is broken into three parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
The third part, Denotational Semantics, introduces a simple
model of semantics, and shows it is compositional, sound, adequate, 
and implies contextual equivalence.
(SF is divided into books, the first two of which have the same names
as the first two parts of PLFA, and cover similar material.)

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an inherently-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.  The new part on
Denotational Semantics also covers material not treated by SF.

@@

There is more material that would be desirable to include in PLFA which was not
due to limits of time, including mutable references, logical relations, System F, and
pure type systems. We'd especially like to include pure type systems as they
provide the readers with a formal model close to the dependent types used in the
book.  Our attempts so far to formalise pure type systems have proved
challenging, to say the least.

@@

There was also a pleasing surprise regarding the structure of the
proofs. While most proofs in both SF and PLFA are carried out by
induction over the evidence that a term is well typed, in SF the
central proof, that substitution preserves types, is carried out by
induction on terms for a technical reason
(the context is extended by a variable binding, and hence not
sufficiently ``generic'' to work well with Coq's induction tactic). In
Agda, we had no trouble formulating the same proof over evidence that
the term is well typed, and didn't even notice SF's description of the
issue until we were done.

@@

\section{Teaching experience}

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  To date, he has taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty two-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA.

@@

The book is published as open-source on Github, under a
Creative Commons CC-BY license.  As the book is an executable
Agda script, it was considered essential that anyone could
download and execute it.

One sign of a successful publication is that it will attract a few
letters from readers who have noticed typos or other problems.  An
unexpected benefit of publishing on Github is that to date forty
readers have sent pull requests.  Most of these fix typos, but a fair
number make more substantial improvements.  Wen wrote a script that
automatically adds the name of any contributor of an accepted pull
request to the list of acknowledgements, sorted by the number of
accepted requests.  Arguably, a different metric, such as total
number of affected lines, might be more appropriate.

There is much left to do!  We hope others may be inspired to join us
and expand the book.

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing how
much more compact it is to avoid raw terms, thanks to David Darais.  For
inspiring our work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.
